Secure Composition of Robust and Optimising Compilers

### 2023-07-10 popl/main.tex
sec 4

\subsubsection{Securing Optimisations}
Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g. as in the work of~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
Consider the compilation passes of LLVM~\cite{lattner2004llvm}.
In LLVM, compiler passes are handled in the \texttt{PassManager} class.
By means of manipulating the internal state of the \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
That way, an order of optimisations can be found by hand that yields the best results.
Take \gls{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls{dce}, which removes dead code by rewriting conditional branches.
Unfortunately, the order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
Examples of this are present in industry:
Consider the Java Virtual Machine bytecode $\trg{JVM BC}$, which is a popular target for programming language designers due to its high performance and relevance in industry.
Compilers for several programming languages have it as their target language, some popular instances are $\src{Java}$ and $\obj{Kotlin}$.
Technically speaking, they both compile to class files and $\obj{Kotlin}$ objects are considered to be the same as $\src{Java}$ objects at that point.
Both languages can be used at the same time in one project~\cite{androidstudio}.
A compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$ code translating to the same target language or intermediate representation performs a kind of {\em upper} composition.
The definition of upper composition is as follows: given a program $\mathtt{p}$, its compiled $\trg{L}$ counterpart is obtained by plugging $\mathtt{p}$ into $\cc{\src{L}}{\trg{L}}$ if $\mathtt{p}\in\src{L}$ or by plugging $\mathtt{p}$ into $\cc{\obj{L}}{\trg{L}}$ if $\mathtt{p}\in\obj{L}$, written as $\cc{\src{L}+\obj{L}}{\trg{L}}$.
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\obj{L}}{\trg{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

Dually, the {\em lower} composition is concerned about compilers that accept the same source but yield different target languages. %and the same theoretical results apply.
Consider two compilers both accepting $\src{LLVM IR}$~\cite{lattner2004llvm} and one of them emits $\trg{x86\_64}$, while the other emits $\trg{ARMv8}$.
It is intuitive that they are in some sense composed in the LLVM framework, but the decision of when to use one over the other is inherently {\em internal} to the formalisation effort of this kind of composition.
For example, the user of this compiler provides an explicit flag that instructs to emit $\trg{x86\_64}$ or the framework itself detects the target platform via heuristics, such as supported instructions.
The %proposed
definition of lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$ is as follows: given a program $\src{p}$, its compiled $\trg{L}$ or $\obj{L}$ program is obtained by performing an internal decision and threading $\src{p}$ into $\cc{\src{L}}{\trg{L}}$ or into $\cc{\src{L}}{\obj{L}}$ depending on that decision.
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\src{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

>>>>

\subsubsection{Securing Optimisations}
Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g. as in the work of~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls{cf} and \gls{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\scriptsize\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls{dce}} (progoptcfdce);
  \caption{Example program where the level of optimisations differ for one pass of applying \gls{cf} and \gls{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls{cf} or \gls{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
%If possible, optimisations are merged to enhance the analysis' result, since even when iterating to a fixpoint the final program may not be as optimised as it could be~\cite{click1995combining}.
Compiler engineers typically try to find an order of optimisations that yields well-optimised programs for either code size~\cite{cooper1999geneticphases} or performance~\cite{kulkarni2006exhaustivephase}.
\Cref{corr:swappable} justifies that any such order of compilation passes is valid with respect to security.
So, given two compilation passes \bul{$\cc[_{1}]{\trg{L}}{\trg{L}}$, $\cc[_{2}]{\trg{L}}{\trg{L}}$, both robustly preserving class $\class[_{1}]$ or $\class[_{2}]$, respectively}, \oul{for any order of their composition the composed compiler robustly preserves the intersection of $\class[_{1}]$ and $\class[_{2}]$}.

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
Examples of this are present in industry:
Consider the Java Virtual Machine bytecode $\trg{JVM BC}$, which is a popular target for programming language designers due to its high performance and relevance in industry.
Compilers for several programming languages have it as their target language, some popular instances are $\src{Java}$ and $\obj{Kotlin}$.
Technically speaking, they both compile to class files and $\obj{Kotlin}$ objects are considered to be the same as $\src{Java}$ objects at that point.
Both languages can be used at the same time in one project~\cite{androidstudio}.
A compiler that accepts both $\src{Java}$ and $\obj{Kotlin}$ code translating to the same target language or intermediate representation performs a kind of {\em upper} composition.
The definition of upper composition is as follows: given a program $\mathtt{p}$, its compiled $\trg{L}$ counterpart is obtained by plugging $\mathtt{p}$ into $\cc{\src{L}}{\trg{L}}$ if $\mathtt{p}\in\src{L}$ or by plugging $\mathtt{p}$ into $\cc{\obj{L}}{\trg{L}}$ if $\mathtt{p}\in\obj{L}$, written as $\cc{\src{L}+\obj{L}}{\trg{L}}$.
\MP{
	put into def env
}
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\obj{L}}{\trg{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.

@@

Dually, the {\em lower} composition is concerned about compilers that accept the same source but yield different target languages. %and the same theoretical results apply.
Consider two compilers both accepting $\src{LLVM IR}$~\cite{lattner2004llvm} and one of them emits $\trg{x86\_64}$, while the other emits $\trg{ARMv8}$.
It is intuitive that they are in some sense composed in the LLVM framework, but the decision of when to use one over the other is inherently {\em internal} to the formalisation effort of this kind of composition.
For example, the user of this compiler provides an explicit flag that instructs to emit $\trg{x86\_64}$ or the framework itself detects the target platform via heuristics, such as supported instructions.
The %proposed
definition of lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$ is as follows: given a program $\src{p}$, its compiled $\trg{L}$ or $\obj{L}$ program is obtained by performing an internal decision and threading $\src{p}$ into $\cc{\src{L}}{\trg{L}}$ or into $\cc{\src{L}}{\obj{L}}$ depending on that decision.
\MP{into def}
Now, given \bul{$\cc{\src{L}}{\trg{L}}$ robustly preserves $\class[_{1}]$} and \rul{$\cc{\src{L}}{\obj{L}}$ robustly preserves $\class[_{2}]$}, it follows that \oul{their lower composition $\cc{\src{L}}{\trg{L}+\obj{L}}$ robustly preserves the intersection of classes $\class[_{1}]$ and $\class[_{2}]$}.
